Efficient Parallelization of $\ell_1$-Minimization for Face Recognition \thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}

### 2011-02-02 paper/main.tex
minor edits to section 2

%Motivate the paper by an overview about the scope and importance of $\ell_1$-minimization.
$\ell_1$-minimization is one of the most powerful tools to be added to the engineer's
toolbox in years.  It has strong theoretical motivation for both for use in recovering sparse
solutions to systems of linear equations, and has been shown to perform well even in the
presence of a combination of sparse and dense error.  Since solutions of linear systems
arise in practically every branch of engineering and there are very few application which
do {\em not} benefit from a robust error function, the applications are far to numerous to
itemize here.  
For many applications of $\ell_1$-minimization solving a system $b=Ax$, $A$ often has a
special structure that can be leveraged to compute portions of $A$ on the fly
as they are needed, speeding up perfomance of the minimization dramatically.
Unfortunately, there are many import applications where this does not hold true.
This paper targets applications where $\bb$ and $A$ contain
image data, and thus must be loaded every time they are used, as is the 
case for many problems in computer vision.
The techniques presented in this paper target these applications, as well
as others where the dictionary is dense and unstructured. $A$ must be loaded every time
it is used, and algorithm performance becomes highly dependent on memory bandwidth, rather
than arithmetic logic.

@@

%The augmented lagrange multiplier (ALM) method utilizes a popular class of convex techniques, the lagrange multiplier, solving the problem:
%\begin{equation}
%F(x) = f(x) + \lambda g(x).
%\end{equation}

%Because the optimal lagrange multiplier values are not known, ALM iterates between estimating the solution and the lagrange multipier values.

%In order to solve the recognition stage, we consider the $l_1$ problem:
%\begin{equation}
%\min_{\x, \e} \|\x\|_1 \quad \subj \quad \bb = A \x + \e.
%\end{equation}
%which can be formulated as
%\begin{equation}
%L_u(x,y) = \|x\|_1 + <\y, b - A\x - \e> + \frac{\mu}2 \| b-A\x-\e \|_2^2
%\end{equation}

%where $\mu > 0$ is a constant that penalizes infeasibility and $y$ is a vector of lagrange multipliers.

We define the following soft-thresholding operator for a
scalar $x$ and a scalar $\alpha \geq 0$:
\begin{equation}
\textup{shrink}(x,\alpha) = \textup{sign}(x)\cdot \max \{|x| - \alpha, 0\},
\end{equation}
In practice, this operator will be applied elementwise to a vector $\x$ with a single $\lambda$,
and will be notated as $\textup{shrink}(\x,\alpha)$.

We summarize the ALM algorithm as Algorithm~\ref{alg:alm}, where $\gamma$ denotes the
largest eigenvalue of the matrix $A^TA$. For the choice of parameter $\mu$, we take the same strategy as
in \cite{YangJ2009-pp} and set $\mu_0 = 2m / \|\bb\|_1$. We set $\rho=1.5$.
\begin{algorithm}[h]
\caption{\bf (Augmented Lagrange Multiplier Method Used in Alignment Inner Loop)}
\begin{algorithmic}[1]
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\blamda_1 = \mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\e_{l+1} \leftarrow \textup{shrink}\left(\bb - A\x_l + \frac{\blamda_k}{\mu_k}, \frac{1}{\mu_k}\right)$;
\STATE $\x_{l+1} \leftarrow (A^\dagger)^T \left(\bb - \e_{l+1} + \frac{\blamda_k}{\mu_k} \right) $;
\ENDWHILE
\STATE $\blamda_{k+1} \leftarrow \blamda_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\STATE $\mu_{k+1} \leftarrow \rho\mu_k$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{algorithmic}
\label{alg:alm}
\end{algorithm}

\subsection{Computational Complexity}
Any discussion of the speed of an algorithm should discuss the comutational complexity of the algorithm.  For the ALM
algorithms considered in this paper, this is rather simple.
The Matrix-vector computions are $O(mn)$, and the vector-vector operations are $O(m)$, 
in terms of both computation cost and memory-bandwidth.  Since A is very tall, 
the computation of $A^\dagger$, the pseudoinverse of $A$, is best computed via $A^\dagger = (A^TA)^-1 A^T$.
The complexity of computing $G = A^T A$ is $O(n*n*m)$.  $G$ is sufficiently small that it can be inverted via
Gauss-Jordan elimination, which has a complexity of $O(n^3)$.  The computation of $A$ finishes with a matrix
multiplication, with cost $O(n^2 m)$.
Fortunately, $A^\dagger$ only needs to be computed once per instance of the $\ell_1$ minimization problem.

Therefore, for problems with large $n$, we expect the computation of the pseudoinverse to dominate.  However,
for the problem sizes needed in face recognition problems, we find that the computation time is dominated
by the repeated computation of matrix-vector and matrix-matrix multiplications in the inner loop.

@@

The ALM algorithm requires many matrix-vector operations, this gives manycore processors such as GPUs a computational advantage for large problem sizes.  
GPUs are comprised of several streaming multiprocessors, which contain 32 floating-point units.  The GPU executes programs as SIMT (Single Instruction Multiple Threads) in groups of 32 threads called warps.  These warps are then grouped into higher levels called thread blocks and grids.  The improvement in speed comes from the memory structure of the GPU.   The GPU contains an inverted memory hierarchy so that the L1 cache stays close to the cores, providing the increase in performance.  It essentially turns the GPU from being memory-limited to computational-limited.
GPU memory also differs from CPU memory in that it requires data transfers over PCI-Express.  The cost of these transfers are amortized by doing large amounts of computation at once, requiring code to be run completely on the GPU.  The GPU can be analogous to a multi-core CPU processor, except with more sophisticated vector processing hardware.  Each streaming multiprocessor is analogous to a single CPU except that it can run multiple threads simultaneously.
All of these differences allow GPUs to perform intensive calculations such as matrix-vector operations at much faster rates than CPUs.

@@

cuda cores, SIMD width, scheduling restrictions, bandwidth
\subsection{Solving a Single $\ell_1$ minimization}

The GPU ALM algorithm is written in CUDA and utilizes the CUBLAS libraries.
CUBLAS are the BLAS libraries consisting of all single precision BLAS calls
which are provided by Nvidia and are optimized for GPU performance. In
addition, we coded kernel (GPU) methods for vector-vector operations to replace
operations that would have taken multiple BLAS calls. The algorithm is similar
to the CPU algorithm except that we had to parallelize the code at an
additional level. The CPU portion of the code serves as the conductor, making
calls to GPU functions and directing the flow of the code while the GPU portion
does all of the actual computation.

\section{Benchmarks on a Single Large $\ell_1$ Problem}

In this section, we present $2$ sets of experiment to benchmark the performance
of the PALM algorithm on CPU and GPU platforms. 
The first set of experiments will compare implementations that solve a single
large instance of an $ell_1$ minimization problem.  
The second set of experiments will compare implementations that must solve
many instances of $\ell_1$ minimization.

To generate a meaningful comparison between different hardware architectures,
in all cases we compare the performance of hardware architectures on a
per-board basis:  For CPU implementations, the benchmark makes use of all of
the cores in as many cpu's are present.  For GPU implementations, the benchmark
makes the best use of the entire GPU chip (most GPU boards have a single GPU
chip).

Unless otherwise specified, all implementations utilize single precision
floating point datatypes.
% and were verified to output results differing by no
%more than $10^{-4}$.

To construct the benchmark for synthetic data, we first created an $m \times n$
matrix $A$ with entries $\pm 1$ each with a $50\%$ chance.  Then we created a
sparse vector $\x_0$ with $k\%$ nonzero random entries.  We also created a
sparse vector $y_k$ with the same number of non-zero entries as $\x$.  Let
$\y_0 = A \x_0 + \y_k$.  We timed the time taken for the ALM algorithm using
$\y_0$, $A$, and other inputs.  The other inputs included, maximum number of
inner/outer loop iterations, and inner/outer loop tolerance level.
% give a discusison of how to construct the benchmark: one paragraph for
% protocol on how to construct synthetic data second P for protocol on how to
% construct the face data

@@

In this section, we benchmark the performance of MATLAB, BLAS MKL, and GPU CUDA
implementations of the Primal ALM algorithm on synthetic data to find at which
point the GPU implementation runs faster than CPU implementation.  To simply
the benchmark, we cap the number of inner iterations at 50 and outer iterations
at 5000 and run the algorithm to convergence.  CPU code was run on an Intel
Dual Socket Quad Core with 8 gigs of DDR3 RAM.  GPU code was run on a GTX480 We
ran the A matrix with different ratios between m:n from 10:1:5 for
m:2000:25:8000 with an x vector of 10 percent sparsity.  Tolerance levels were
empirically seen to allow best performance with the inner loop tolerance set at
one order of magnitude lower than the outer loop tolerance.

@@

\section{Benchmarks on Solving Many $\ell_1$ Problems Concurrently}

%This section is about Drew's kernels for solving multiple $\ell_1$ minimizations in parllel

%How to efficiently perform an $\ell_1$ minimization on a single core, quote performance on a large batch.

%How to solve multiple instances in smaller batches without losing concurrency, using CUDA streams.

%...Mention effect of compiler verion on streams implementation

>>>>

%Motivate the paper by an overview about the scope and importance of $\ell_1$-minimization.
$\ell_1$-minimization is one of the most powerful tools to be added to the engineer's
toolbox in years.  It has strong theoretical motivation for both for use in recovering sparse
solutions to systems of linear equations, and has been shown to perform well even in the
presence of a combination of sparse and dense error.  Since solutions of linear systems
arise in practically every branch of engineering and there are very few application which
do {\em not} benefit from a robust error function, the applications are far to numerous to
itemize here.  
For many applications of $\ell_1$-minimization solving a system $b=Ax$, $A$ often has a
special structure that can be leveraged to compute portions of $A$ on the fly
as they are needed, speeding up perfomance of the minimization dramatically.
Unfortunately, there are many import applications where this does not hold true.
This paper targets applications where $\bb$ and $A$ contain
image data, and thus must be loaded every time they are used, as is the 
case for many problems in computer vision.
The techniques presented in this paper target these applications, as well
as others where the dictionary is dense and unstructured. $A$ must be loaded every time
it is used, and algorithm performance becomes highly dependent on memory bandwidth, rather
that arithmetic logic.

@@

In this section, we describe the general version ALM algorithm and then describe how it can be adapted to solve our problem.

The augmented lagrange multiplier (ALM) method utilizes a popular class of convex techniques, the lagrange multiplier, solving the problem:
\begin{equation}
F(x) = f(x) + \lambda g(x).
\end{equation}
which can be formulated as an augmented Langrange function given by
\begin{equation}
L_u(x,y) = \|x\|_1 + <\y, b - A\x - \e> + \frac{\mu}2 \| b-A\x-\e \|_2^2
\end{equation}

where $\mu > 0$ is a constant that penalizes infeasibility and $y$ is a vector of lagrange multipliers.

For an optimal $y*$ which satisfies the second-order sufficiency conditions, for a sufficiently large $mu$, it can be shown that
\begin{equation}
x* = \arg min L_u(x,y*)
\end{equation}

However, because $y*$ is generally unknown, ALM methods simultaneously estimate the optimal solution and Lagrange multipliers iteratively in the following procedure.

We also need to define the following soft-thresholding operator for a
scalar $x$ and a scalar $\alpha \geq 0$:
\begin{equation}
\textup{shrink}(x,\alpha) = \textup{sign}(x)\cdot \max \{|x| - \alpha, 0\},
\end{equation}
In practice, this operator will be applied elementwise to a vector $\x$ with a single $\lambda$,
and will be notated as $\textup{shrink}(\x,\alpha)$.

We summarize the entire ALM
algorithm as Algorithm~\ref{alg:alm}, where $\gamma$ denotes the
largest eigenvalue of the matrix $A^TA$. For the choice of parameter $\mu$, we take the same strategy as
in \cite{YangJ2009-pp} and set $\mu_0 = 2m / \|\bb\|_1$. We set $\rho=1.5$.
\begin{algorithm}[h]
\caption{\bf (Augmented Lagrange Multiplier Method Used in Alignment Inner Loop)}
\begin{algorithmic}[1]
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\blamda_1 = \mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\e_{l+1} \leftarrow \textup{shrink}\left(\bb - A\x_l + \frac{\blamda_k}{\mu_k}, \frac{1}{\mu_k}\right)$;
\STATE $\x_{l+1} \leftarrow (A^\dagger)^T \left(\bb - \e_{l+1} + \frac{\blamda_k}{\mu_k} \right) $;
\ENDWHILE
\STATE $\blamda_{k+1} \leftarrow \blamda_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\STATE $\mu_{k+1} \leftarrow \rho\mu_k$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{algorithmic}
\label{alg:alm}
\end{algorithm}

%Show Pseudocode for one variant
%\subsection{Computational Complexity}
The Matrix-vector computions are O($mn$), and the vector-vector operations are O($m$), 
in terms of both computation cost and memory-bandwidth.  As A'A can be precomputed once, assuming the iterative solver always takes a constant number of iterations, the PALM algorithm has a computational complexity of O($mn+n^2$), where n $<$ m.

% Complexity of entire algorithm

\subsection{Adapations}

In this section, we describe how to adapt the general ALM algorithm for the alignment and recognition problem.  Refer to book...

In order to solve the alignment stage before recognition, we consider the $l_1$ problem: 
\begin{equation} 
\min_{\x, \e} \|\e\|_1 \quad \subj \quad \bb = A \x + \e.
\end{equation}
which is solved in a similar fashion.

@@

The ALM algorithm requires many matrix-vector operations, this gives manycore processors such as GPUs a computational advantage for large problem sizes.  
GPUs are comprised of several streaming multiprocessors (SMP) (16 for the Nvidia Fermi), which each contain 32 floating-point units.  Each SMP contains its own L1-cache and shared memory and all SMPs share a common L2-cache.  The GPU executes programs as SIMT (Single Instruction Multiple Threads) in groups of 32 threads called warps.  These warps are then grouped into higher levels called thread blocks and grids.
%The GPU contains an inverted memory hierarchy so that the L1 cache stays close to the cores, providing the increase in performance.  It essentially turns the GPU from being memory-limited to computational-limited.
GPU memory also differs from CPU memory in that it requires data transfers over PCI-Express, which have limited bandwidth.  The cost of these transfers are amortized by doing large amounts of computation at once, requiring code to be run completely on the GPU.  The GPU can be analogous to a multi-core CPU processor, except with more sophisticated vector processing hardware.  Each thread block is a virtualized multiprocessor, which each thread as a virtualized scalar processor.  All of these differences allow GPUs to perform intensive calculations such as matrix-vector operations at much faster rates than CPUs.

%Each SMP is analogous to a single CPU except that it can run multiple threads simultaneously.

@@

%cuda cores, SIMD width, scheduling restrictions, bandwidth
\subsection{Solving a Single $\ell_1$ minimization}

The GPU ALM algorithm is written in CUDA and utilizes the CUBLAS libraries.
CUBLAS are the BLAS libraries consisting of all single precision BLAS calls
which are provided by Nvidia and are optimized for GPU performance. In
addition, we coded kernel (GPU) methods for vector-vector operations to replace
operations that would have taken multiple BLAS calls. Because data transfer from the RAM to GPU suffers bandwidth limitations, we designed our implementation to transfer the A matrix and measurement vector at once.  The algorithm is similar
to the CPU algorithm except that we had to parallelize the code at an
additional level. The CPU portion of the code serves as the conductor, making
calls to GPU functions and directing the flow of the code while the GPU keeps all of the data and runs the computation.

\subsection{Solving multiple $\ell_1$ minimizations in parallel}
This section is about Drew's kernels for solving multiple $\ell_1$ minimizations in parllel

How to efficiently perform an $\ell_1$ minimization on a single core, quote performance on a large batch.

How to solve multiple instances in smaller batches without losing concurrency, using CUDA streams.

...Mention effect of compiler verion on streams implementation

\section{Benchmarks}

In this section, we present INSERTNUMBERHERE sets of experiment to benchmark the performance
of the PALM algorithm on CPU and GPU platforms. To generate a meaningful
comparison between different hardware architectures, we compare on a per-board
basis:  For CPU implementations, the benchmark makes use of as all of the cores
in as many cpu's are present.  For GPU implementations, the benchmark makes the
best use of the entire GPU chip (most GPU boards have a single GPU chip). Both
implementations utilized single precision floating point datatypes and were
verified to output results differing by no more than $10^{-4}$.

To construct the benchmark for synthetic data, we first created an $m \times n$
matrix $A$ with entries $\pm 1$ each with a $50\%$ chance.  Then we created a
sparse vector $\x_0$ with $10\%$ nonzero random entries.  We also created a
sparse vector $y_k$ with the same number of non-zero entries as $\x$.  Let
$\y_0 = A \x_0 + \y_k$.  We timed the time taken for the ALM algorithm using
$\y_0$, $A$, and other inputs.  The other inputs included, maximum number of
inner/outer loop iterations, and inner/outer loop tolerance level.
% give a discusison of how to construct the benchmark: one paragraph for
% protocol on how to construct synthetic data second P for protocol on how to
% construct the face data

@@

In this section, we benchmark the performance of BLAS MKL and GPU CUDA
implementations of the Primal ALM algorithm on synthetic data to find at which
point the GPU implementation runs faster than CPU implementation.  To simply
the benchmark, we cap the number of inner iterations at 50 and outer iterations
at 5000 and run the algorithm to convergence.  CPU code was run on an Intel
Dual Socket Quad Core with 8 gigs of DDR3 RAM.  GPU code was run on a GTX480 We
ran the A matrix with different ratios between m:n from 10:1:5 for
m:2000:25:8000 with an x vector of 10 percent sparsity.  Tolerance levels were
empirically seen to allow best performance with the inner loop tolerance set at
one order of magnitude lower than the outer loop tolerance.

@@
